A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-03-19 03-semantics.tex
Typos levantados pelos revisores

At the beginning of the reaction, an external event is emitted, which
will trigger rule \R{push}, which will immediately raise the stack level
to~1.
At the end of the reaction, the program will block or terminate and
successive applications of
rule~\R{pop} will eventually lead to a description containing this
same program at stack level~0.

@@

\begin{figure}[h]
\small
\begin{gather*}
  \boxed{
    \begin{align*}
      %%
      %%-
      \intertext{\llap{(i)~}Function~$\bcast$:}
      %%-
      %%
      \bcast(\ceu{\AwaitExt(e)},e)
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \bcast(\ceu{\AwaitInt(e)},e)
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \bcast(\ceu{\Every{e}\ {p}},e)
      &=\ceu{p;\,\Every{e}\ {p}}\\[-1\jot]
      %%
      \bcast(\ceu{\CanRun(n)},e)
      &=\ceu{\CanRun(n)}\\[-1\jot]
      %%
      \bcast(\ceu{\Fin{p}},e)
      &=\ceu{\Fin{p}}\\[-1\jot]
      %%
      \bcast(\ceu{p;\,q})
      &=\ceu{\bcast(p,e);\,q}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtLoop{q}},e)
      &=\ceu{\bcast(p,e)\AtLoop{q}}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtAnd{q}},e)
      &=\ceu{{\bcast(p,e)}\AtAnd{\bcast(q,e)}}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtOr{q}},e)
      &=\ceu{{\bcast(p,e)}\AtOr{\bcast(q,e)}}\\[-1\jot]
      %%
      bcast(\_,e)
      &=\_\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
      \\[1\jot]
      %%
      %%-
      \intertext{\llap{(ii)~}Predicate~$\isblocked$:}
      %%-
      %%
      \isblocked(\ceu{\AwaitExt(e)},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\AwaitInt(e)},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\Every{e}\ {p}},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\CanRun(m)},n)
      &=(n>m)\\[-1\jot]
      %%
      \isblocked(\ceu{\Fin{p}},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{p;\,q},n)
      &=\isblocked(p,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtLoop{q}},n)
      &=\isblocked(p,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtAnd{q}},n)
      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtOr{q}},n)
      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
      %%
      \isblocked(\_,n)
      &=\mathit{false}\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
        \\[1\jot]
      %%
      %%-
      \intertext{\llap{(iii)~}Function~$\clear$:}
      %%-
      %%
      \clear(\ceu{\AwaitExt(e)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\AwaitInt(e)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\Every{e}\ p})
      %%
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\CanRun(n)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\Fin{p}})
      &=p\\[-1\jot]
      %%
      \clear(\ceu{p;\,q})
      &=\clear(p)\\[-1\jot]
      %%
      \clear(\ceu{p\AtLoop{q}})
      &=\clear(p)\\[-1\jot]
      %%
      \clear(\ceu{p\AtAnd{q}})
      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
      %%
      \clear(\ceu{p\AtOr{q}})
      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
      %%
      \clear(\_)
      &=\xi\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
    \end{align*}}
\end{gather*}
\vskip-2\belowdisplayskip
\caption{%
  (i)~Function~$\bcast$ awakes awaiting trails matching the event by
  converting~$\ceu{\protect\AwaitExt}$ and~$\ceu{\protect\AwaitInt}$
  to~$\ceu{\protect\Nop}$, and by unwinding $\ceu{\protect\Every}$
  expressions.
  %%
  \space(ii)~Predicate~$\isblocked$ is true only if all branches in parallel
  are blocked waiting for events, finalization clauses, or certain
  stack levels.
  %%
  \space(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ expressions in
  parallel and put their bodies in sequence.
  %%
  In~(i), (ii), and~(iii),~``$\_$'' denotes the omitted cases and~``$\xi$''
  denotes the empty string.
  %%
}
\label{fig.bcast}
\label{fig.isblocked}
\label{fig.clear}
\end{figure}

@@

The next two lemmas establish the determinism of a single application
of~$\out$ and~$\nst$.  Lemma~\ref{lem.x.det-out} follows from a simple
inspection of rules~\R{push} and~\R{pop}.  The proof of
Lemma~\ref{lem.x.det-out}, however, requires an induction on the structure
of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
used in the proof of Theorem~\ref{thm.x.det}, the main result of this
section.  Theorem~\ref{thm.x.det} establishes that any given number of
applications of~$\trans$ starting from the same input description will
always lead to the same output description.

@@

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
The idea of the proof is that a sufficiently large sequence of~$\nst$
and~$\out$ transitions eventually decrease the rank of the current
description until an irreducible description is reached.  This irreducible
description is the final result of the reaction.
%%
\begin{theorem}[Termination]\label{thm.x.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.x.rank-nst-*},
    $\<i,j>=\<0,0>\ge\rank(\delta')$,
    %%
    % \[
    %   \<i,j>=\<0,0>\ge\rank(\delta')\,,
    % \]
    %%
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases depending on whether or not $\delta$~is
    nested-irreducible.
    \begin{case}
      \label{thm.x.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
      that $j>0$.  Then there are two subcases.
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nonnil}
        $e\ne\nil$.
        %%
        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
        \begin{align*}
          \rank(\delta)=\rank(\delta_1')=\<i,j>
                                        \ge\rank(\delta')=\<i',j'>\,.
        \end{align*}
        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
        was consumed by the nested transitions.  Thus,
        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
        $\delta\trans[*]\delta''_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nil}
        $e=\nil$.
        %%
        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta'$.  By
        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
        Therefore, $\delta\trans[*]\delta''_\#$.
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
    \end{case}
  \end{induction}
\end{proof}

>>>>

At the beginning of a reaction, an external event is emitted, which
will trigger rule \R{push}, which will immediately raise the stack level
to~1.
At the end of the reaction, the program will block or terminate and
successive applications of
rule~\R{pop} will eventually lead to a description containing this
same program at stack level~0.

@@

\begin{figure}[h]
\small
\begin{gather*}
  \boxed{
    \begin{align*}
      %%
      %%-
      \intertext{\llap{(i)~}Function~$\bcast$:}
      %%-
      %%
      \bcast(\ceu{\AwaitExt(e)},e)
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \bcast(\ceu{\AwaitInt(e)},e)
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \bcast(\ceu{\Every{e}\ {p}},e)
      &=\ceu{p;\,\Every{e}\ {p}}\\[-1\jot]
      %%
      \bcast(\ceu{\CanRun(n)},e)
      &=\ceu{\CanRun(n)}\\[-1\jot]
      %%
      \bcast(\ceu{\Fin{p}},e)
      &=\ceu{\Fin{p}}\\[-1\jot]
      %%
      \bcast(\ceu{p;\,q},e)
      &=\ceu{\bcast(p,e);\,q}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtLoop{q}},e)
      &=\ceu{\bcast(p,e)\AtLoop{q}}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtAnd{q}},e)
      &=\ceu{{\bcast(p,e)}\AtAnd{\bcast(q,e)}}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtOr{q}},e)
      &=\ceu{{\bcast(p,e)}\AtOr{\bcast(q,e)}}\\[-1\jot]
      %%
      bcast(\_,e)
      &=\_\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
      \\[1\jot]
      %%
      %%-
      \intertext{\llap{(ii)~}Predicate~$\isblocked$:}
      %%-
      %%
      \isblocked(\ceu{\AwaitExt(e)},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\AwaitInt(e)},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\Every{e}\ {p}},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\CanRun(m)},n)
      &=(n>m)\\[-1\jot]
      %%
      \isblocked(\ceu{\Fin{p}},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{p;\,q},n)
      &=\isblocked(p,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtLoop{q}},n)
      &=\isblocked(p,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtAnd{q}},n)
      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtOr{q}},n)
      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
      %%
      \isblocked(\_,n)
      &=\mathit{false}\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
        \\[1\jot]
      %%
      %%-
      \intertext{\llap{(iii)~}Function~$\clear$:}
      %%-
      %%
      \clear(\ceu{\AwaitExt(e)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\AwaitInt(e)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\Every{e}\ p})
      %%
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\CanRun(n)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\Fin{p}})
      &=p\\[-1\jot]
      %%
      \clear(\ceu{p;\,q})
      &=\clear(p)\\[-1\jot]
      %%
      \clear(\ceu{p\AtLoop{q}})
      &=\clear(p)\\[-1\jot]
      %%
      \clear(\ceu{p\AtAnd{q}})
      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
      %%
      \clear(\ceu{p\AtOr{q}})
      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
      %%
      \clear(\_)
      &=\xi\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
    \end{align*}}
\end{gather*}
\vskip-2\belowdisplayskip
\caption{%
  (i)~Function~$\bcast$ awakes awaiting trails matching the event by
  converting~$\ceu{\protect\AwaitExt}$ and~$\ceu{\protect\AwaitInt}$
  to~$\ceu{\protect\Nop}$, and by unwinding $\ceu{\protect\Every}$
  expressions.
  %%
  \space(ii)~Predicate~$\isblocked$ is true only if all branches in parallel
  are blocked waiting for events, finalization clauses, or certain
  stack levels.
  %%
  \space(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ expressions in
  parallel and put their bodies in sequence.
  %%
  In~(i), (ii), and~(iii),~``$\_$'' denotes the omitted cases and~``$\xi$''
  denotes the empty string.
  %%
}
\label{fig.bcast}
\label{fig.isblocked}
\label{fig.clear}
\end{figure}

@@

The next two lemmas establish the determinism of a single application
of~$\out$ and~$\nst$.  Lemma~\ref{lem.x.det-out} follows from a simple
inspection of rules~\R{push} and~\R{pop}.  The proof of
Lemma~\ref{lem.x.det-nst}, however, requires an induction on the structure
of the derivation trees produced by the rules for~$\nst$.  Both lemmas are
used in the proof of Theorem~\ref{thm.x.det}, the main result of this
section.  Theorem~\ref{thm.x.det} establishes that any given number of
applications of~$\trans$ starting from the same input description will
always lead to the same output description.

@@

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
The idea of the proof is that a sufficiently large sequence of~$\nst$
and~$\out$ transitions eventually decrease the rank of the current
description until an irreducible description is reached.  This irreducible
description is the final result of the reaction.
%%
\begin{theorem}[Termination]\label{thm.x.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.x.rank-nst-*},
    $\<i,j>=\<0,0>\ge\rank(\delta')$,
    %%
    % \[
    %   \<i,j>=\<0,0>\ge\rank(\delta')\,,
    % \]
    %%
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases depending on whether or not $\delta$~is
    nested-irreducible.
    \begin{case}
      \label{thm.x.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
      that $j>0$.  Then there are two subcases.
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nonnil}
        $e\ne\nil$.
        %%
        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
        \begin{align*}
          \rank(\delta)=\rank(\delta_1')=\<i,j>
                                        \ge\rank(\delta')=\<i',j'>\,.
        \end{align*}
        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
        was consumed by the nested transitions.  Thus,
        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
        $\delta\trans[*]\delta''_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nil}
        $e=\nil$.
        %%
        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta'$.  By
        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
        Therefore, $\delta\trans[*]\delta''_\#$.
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorem~\ref{thm.x.term-nst-*}
      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
    \end{case}
  \end{induction}
\end{proof}

### 2018-03-19 index.tex
Typos levantados pelos revisores

\begin{abstract}
\CEU is a synchronous programming language for embedded soft real-time systems.
%
It focus on control-flow safety features, such as safe shared-memory
concurrency and safe abortion of lines of execution, while enforcing
memory-bounded, deterministic, and terminating reactions to the environment.
%
In this work, we present a small-step structural operational semantics for
\CEU and a proof that reactions have the properties enumerated above:
%
that for a given arbitrary timeline of input events, multiple executions of the
same program always react in bounded time and arrive at the same final finite
memory state.
%
%We also discuss some equivalence results and the relation between the proposed
%semantics and its actual implementation.
\end{abstract}

>>>>

\begin{abstract}
\CEU is a synchronous programming language for embedded soft real-time systems.
%
It focuses on control-flow safety features, such as safe shared-memory
concurrency and safe abortion of lines of execution, while enforcing
memory-bounded, deterministic, and terminating reactions to the environment.
%
In this work, we present a small-step structural operational semantics for
\CEU and a proof that reactions have the properties enumerated above:
%
that for a given arbitrary timeline of input events, multiple executions of the
same program always react in bounded time and arrive at the same final finite
memory state.
%
%We also discuss some equivalence results and the relation between the proposed
%semantics and its actual implementation.
\end{abstract}
